---
title: "331 Final Project"
subtitle: "Can Money Buy Happiness?"
author: "Justin Koida, Abigayle Mercer, Sammy Paykel"
date: "2/26/2025"
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    embed-resources: true
    toc: true
    theme: lux
editor: 
  source: true
execute:
  echo: true
  error: false
  output: false
  message: false
  warning: false
---

### 1.0 // Introduction

##### Data description:

Country: Country

Year: Year

Poverty: The percentage of the population in poverty (0-100). Poverty here is defined as living off of less than \$3.65 per day. This is not adjusted for inflation.

Happiness Score: Happiness as a percent indicated by WHR. The World Happiness Report (WHR) is scored from the national average response to the questions of life evaluations.

Observational Unit: Country and Year

##### Hypothesis:

Happiness and Poverty levels are linearly correlated. So we hypothesize that money can buy happiness.

##### Data cleaning process and decision:

We chose to drop all observations with na values, we may go back later and choose to do some data imputation to keep more data. However, we found ththat dropping na resulted in a loss of \~800 data points from the original 3200.

### 1.1 // Data

```{r setup}
library(tidyverse)
library(DT)

pop <- read_csv("data/pop.csv")
hap <- read_csv("data/hapiscore_whr.csv")
pov <- read_csv("data/gm_365pov_rate.csv")
```

### 1.2, 1.3, 1.4, 1.5 // Data Cleaning and Joining

```{r data-cleaning}
#| output: true

clean_hap <- hap |>
  pivot_longer(cols = `2005`:`2023`,
               names_to = "Year",
               values_to = "Happiness") |>
  rename(Country = country) |>
  drop_na()

clean_pov <- pov |>
  select(`2005`:`2023`, country) |>
  pivot_longer(cols = `2005`:`2023`,
               names_to = "Year",
               values_to = "Poverty") |>
  rename(Country = country) |>
  drop_na()

joined_pov_hap <- clean_hap |>
  inner_join(clean_pov, by = join_by(Country == Country, Year == Year))

datatable(joined_pov_hap, 
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 24px;',
            "Poverty and Happiness Data, Joined by Country and Year"
          ), 
          options = list(pageLength = 10)
          )
```

### 2.0 // Linear Regression

```{r}
#| label: trend-over-time
#| output: true

# Create a new variable: Happiness/Poverty ratio
joined_pov_hap <- joined_pov_hap |>
  mutate(Hap_Pov_Ratio = Happiness / Poverty)

# Plot boxplots of the ratio over time
ggplot(joined_pov_hap, aes(x = as.factor(Year), y = Hap_Pov_Ratio)) +
  geom_boxplot(fill = "lightblue", alpha = 0.6) +  # Boxplots without outliers
  geom_jitter(width = 0.2, alpha = 0.3, color = "blue") +  # Jitter for individual points
  
  labs(title = "Distribution of Happiness/Poverty Ratio Over Time",
       x = "Year",
       y = "Happiness / Poverty Ratio") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

```

```{r}
#| label: happiness-poverty-ratio-over-time
#| output: true

pov_hap_avg <- joined_pov_hap |>
  group_by(Country) |>
  summarize(Avg_Happiness = mean(Happiness, na.rm = TRUE),
            Avg_Poverty = mean(Poverty, na.rm = TRUE), .groups = "drop")

# Scatterplot with one observation per country
ggplot(pov_hap_avg, aes(x = Avg_Poverty, y = Avg_Happiness)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Averaged Relationship Between Happiness and Poverty",
       x = "Average Poverty Rate (%)",
       y = "Average Happiness Score (%)") +
  theme_minimal()
```

##### 2.2 // Linear Regression

Linear Regression is a supervised machine learning technique to learn the relationship between a dependent variable and one or more independent variables. Linear Regression assumes that there is some sort of linear relationship between the two variables. Linear Regression attempts to calculate the best Fit line, where the slope indicates the strength of the relationship between the two variables.

source: <https://www.geeksforgeeks.org/ml-linear-regression/>

```{r 2.2}
#| output: true
model <- lm(Happiness ~ Poverty, data = joined_pov_hap) #https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/predict.lm
summary(model)
```

##### 2.3 // Model Fit

```{r 2.3}
#| output: true
# print(tibble(predict(model)))

pre_variance = joined_pov_hap |>
  select(Happiness) |>
  mutate(predicted_happiness = predict(model)) |> # https://www.math.ucla.edu/~anderson/rw1001/library/base/html/predict.lm.html#:~:text=lm%20produces%20predicted%20values%2C%20obtained,of%20the%20predictions%20are%20calculated.
  mutate(residual = Happiness - predicted_happiness)

variances = pre_variance |> # https://www.marsja.se/variance-in-r-how-to-find-calculate/
  mutate(happiness_variance = var(Happiness), 
         predicted_variance = var(predicted_happiness),
         residual_variance = var(residual)) |>
  select(happiness_variance, predicted_variance, residual_variance) |>
  head(1)

knitr::kable(variances)
```

###### Model Fit Analysis

The linear regression model predicting Happiness from Poverty produced an ( R\^2 ) value of 0.5536, meaning it explains \~55.4% of the variation in happiness.

###### Interpretation:

-   The negative coefficient for Poverty (-0.239) suggests a moderate inverse relationship: as poverty increases, happiness tends to decrease.
-   The model is statistically significant (( p \< 2.2 \* 10\^{-16} )).
-   However, about 45% of the variance remains unexplained, indicating other factors (e.g., GDP, social policies) likely influence happiness.
