---
title: "331 Final Project - Project Proposal"
subtitle: "Can Money Buy Happiness?"
author: "Justin Koida, Abigayle Mercer, Sammy Paykel"
date: "2/26/2025"
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    embed-resources: true
    toc: true
    theme: lux
editor: 
  source: true
execute:
  echo: true
  error: false
  output: false
  message: false
  warning: false
---

### 1.0 // Introduction

##### Data description: 

Country: Country 

Year: Year

Poverty: 
The percentage of the population in poverty (0-100). Poverty here is defined as living off of less than $3.65 per day. This is not adjusted for inflation. 

Happiness Score: 
Happiness as a percent indicated by WHR. The World Happiness Report (WHR) is scored from the national average response to the questions of life evaluations. 

Observational Unit: Country and Year

##### Hypothesis: 
Happiness and Poverty levels are linearly correlated. So we hypothesize that money can buy happiness. 

##### Data cleaning process and decision: 
We chose to drop all observations with na values, we may go back later and choose to do some data imputation to keep more data. However, we found ththat dropping na resulted in a loss of ~800 data points from the original 3200. 

### 1.1 // Data
```{r setup}
library(tidyverse)
library(DT)

pop <- read_csv("data/pop.csv")
hap <- read_csv("data/hapiscore_whr.csv")
pov <- read_csv("data/gm_365pov_rate.csv")
```

### 1.2, 1.3, 1.4, 1.5 // Data Cleaning and Joining
```{r data-cleaning}
#| output: true

clean_hap <- hap |>
  pivot_longer(cols = `2005`:`2023`,
               names_to = "Year",
               values_to = "Happiness") |>
  rename(Country = country) |>
  drop_na()

clean_pov <- pov |>
  select(`2005`:`2023`, country) |>
  pivot_longer(cols = `2005`:`2023`,
               names_to = "Year",
               values_to = "Poverty") |>
  rename(Country = country) |>
  drop_na()

joined_pov_hap <- clean_hap |>
  inner_join(clean_pov, by = join_by(Country == Country, Year == Year))

datatable(joined_pov_hap, 
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 24px;',
            "Poverty and Happiness Data, Joined by Country and Year"
          ), 
          options = list(pageLength = 10)
          )
```

